---
title: "by_row(): An alternative to map() or rowwise()"
author: "Matthew Kay"
date: "3/3/2021"
output: github_document
---

This is a short proposal for a new alternative to `rowwise()` or `map()` for rowwise operations on data frames inside mutate.

```{r message=FALSE}
library(tidyverse)
library(rlang)
```

It's useful to be able to perform rowwise operations on data frames. As a trivial
example (obviously these shouldn't be list columns in practice, but ignore that for now), this fails:

```{r error=TRUE}
tibble(a = list(10, 11), b = list(3, 4)) %>%
  mutate(c = a + 2 * b)
```

One way to do rowwise operations is to use `lapply()` or `map()` inside `mutate()`, but
this requires you to pass through all the variables you want to use in an awkward way:

```{r}
tibble(a = list(10, 11), b = list(3, 4)) %>%
  mutate(c = map2_dbl(a, b, ~ .x + 2 * .y))
```

With `rowwise()` we can avoid managing all these variables:

```{r}
tibble(a = list(10, 11), b = list(3, 4)) %>%
  rowwise() %>%
  mutate(c = a + 2 * b)
```

But `rowwise()` has significant drawbacks. It changes the meaning of 
`mutate()` in a way that I have found confuses students---and, frankly, 
experts (if I can be called an expert). If you don't know the data frame 
is in a `rowwise()` state, you can write `mutate()` calls that will raise
errors that don't make sense. Or if you insert a `rowwise()` call into
a pipe for a `mutate()` and forget to ungroup the data frame afterwards,
it can break existing `mutate()` calls downstream. 

The root of the problem, I think, is that `rowwise()` forces people to keep track
of state in a way that humans aren't great at: we have to remember if a data frame
is in "rowwise" or "not rowwise" state for every `mutate()` call, and the
call that puts you in `rowwise()` mode can be arbitrarily far from the corresponding
mutate call. In human-computer interaction we
call errors due to thinking a system is in a different state a [mode error](https://en.wikipedia.org/wiki/Mode_(user_interface)#Mode_errors); it's
the reason why people hate the CapsLock key---we're not good at keeping track
of modes!

I think it would be better to explicitly "mark out" expressions intended to be
executed in a row-wise manner (similar to how `map()` works), but not have to
deal with managing variables (similar to how `rowwise()` works). The best
of both worlds!

My modest proposal is a function like this (which no doubt needs work to handle
some NSE corner cases I haven't thought of):

```{r}
by_row = function(expr) {
  .expr = enquo(expr)
  .data = cur_data_all()
  .env = parent.frame()

  simplify(lapply(seq_len(nrow(.data)), function(i) {
    eval_tidy(.expr, data = lapply(.data[i,], `[[`, 1), env = .env)
  }))
}
```

Which can be used like this:

```{r}
tibble(a = list(10, 11), b = list(3, 4)) %>%
  mutate(c = by_row(a + 2 * b))
```

The other bonus here is that this can be intermixed with non-rowwise calls in the
same `mutate()` call. This is important, because *rowwise operations are slow and you should
try to avoid them at all costs*! Putting the entire dataframe into a rowwise mode
to perform these operations encourages people to do other operations rowwise that
they don't need to do.
